---
title: "Reproducible Research: Storm Severity Analysis"
output: 
  html_document:
    keep_md: true
---

## Questions
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Project Setup

The libraries used in this analysis are:
```{r setup}
library(reshape2)
library(data.table)
library(plyr); library(dplyr)
library(ggplot2)
library(stringr)
```

## Loading and preprocessing

The data comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. An intermediate download location for this class is hosted on cloudfront:

```{r download, cache=TRUE}
setwd("~/../datascience/storm-severity")
dir.create("data")

dlURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filepath <- "data/StormData.csv.bz2"

# set browser to download file through IE protocol (for https)
setInternet2(use = TRUE)
download.file(dlURL, destfile = filepath, mode="wb")
```

```{r load, cache=TRUE}
stormdata <- read.csv(filepath, stringsAsFactors=FALSE);
```

To answer the two questions posed earlier we only need to care about the columns for `EVTYPE`, `FATALITIES`, `INJURIES`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG`, `CROPDMGEXP`, and `REMARKS`. 
```{r preprocess}
stormdat <- data.table(select(stormdata, EVTYPE, FATALITIES:CROPDMGEXP, REMARKS))
stormdat <- mutate(stormdat, EVTYPE=factor(EVTYPE), 
                   PROPDMGEXP=factor(PROPDMGEXP), 
                   CROPDMGEXP=factor(CROPDMGEXP))
dim(stormdat)
sapply(stormdat[1,], class)
head(stormdat)
```

## Any missing Values? 
```{r NAcount}
NA.count <- !complete.cases(select(stormdat, FATALITIES:PROPDMG, CROPDMG))
```
There are `r sum(NA.count)` rows with NA values.

## Managing EVTYPE variable
There are `r length(unique(stormdata$EVTYPE))` different event types in original `EVTYPE` data, which is much larger than the defined list in the [NOAA instruction sheet](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). Let's see if this can be cleaned up. 

### Reducing data set for ease of manipulation
This is a fairly large data set, especially considering the need to do some fairly manual cleaning of the EVTYPE data, as discussed above. First we ignore data which had no adverse impact.  

```{r data.focus}
# filter for adverse effects > 0
stormdat <- filter(stormdat, (FATALITIES+INJURIES+PROPDMG+CROPDMG) > 0)
dim(stormdat)
length(unique(stormdat$EVTYPE))
```

### Basic string cleanup
```{r evtype.cleanup}
# clean up variations
stormdat <- mutate(stormdat, 
                   EVTYPE=str_trim(EVTYPE),
                   EVTYPE=tolower(EVTYPE),
                   EVTYPE=gsub("\\.$", "", EVTYPE),
                   EVTYPE=gsub("/ ", "/", EVTYPE))

evtypes <- unique(stormdat$EVTYPE)
length(evtypes)
```
That helped a little, but there's still too much data.

Taking a look at values of `EVTYPE`:
```{r evtype.values}
head(evtypes, 15)
```
shows several spelling variations for the entry `THUNDERSTORM WINDS`. Also various hurricanes are named and labeled with variations such as `Hurricane Opal`, `Hurricane Erin`, `Hurricane Opal/High Winds`. 

The "real" event types are extracted from the table of contents of the NOAA instruction sheet.
```{r evtype.definition}
# get event types from event types text file
evtypes.raw <- readLines("data/event types.txt", warn=F)

# extract 1st-level event types, excluding "...", page numbers and other data
#   "7.2 Words, (with-stuff/other stuff) "
evtypes.def <- str_extract_all(evtypes.raw, 
                               "^[0-9]+\\.[0-9]+\\s[A-Za-z\\s,/()-]{2,}")

# remove empty rows
evtypes.def <- unlist(evtypes.def[sapply(evtypes.def, length) > 0])

# remove leading numbers and trailing (<char>)
evtypes.def <- str_sub(evtypes.def, 
                       str_locate(evtypes.def, "[0-9] ")[,2]+1, 
                       str_locate(evtypes.def, " \\([A-Z]\\)")[,1]-1)

evtypes.def <- tolower(evtypes.def)
```

While a better approach would be to merge the evtypes.def and evtypes variables via code, time constraints required simply fixing a few spelling variations. Ideally all of the remaining `r length(evtypes)` event types in the data would be mapped to the `r length(evtypes.def)` defined event types defined from the NOAA documentation above. 
```{r evtype.mapping}

# organize event type data to match evtype.def
evtypes <- evtypes[order(evtypes)]
evtypes <- data.table(type.raw=evtypes, type.mapped=evtypes)

# fix common spelling errors for wind, flood, freeze
fixtext <- function(strings, patt, repl) sapply(strings, gsub, 
                                                pattern=patt,
                                                replacement=repl)
pattern.flood <- "\\bflood(ing|in|ed|s)?\\b"
pattern.freeze <- "\\b(freez(e|ing|in|ed|s)?|frozen)\\b"
pattern.wind <- "\\bwi(n|nd|nds|ds|s)?\\b"
pattern.thunderstormwind <- "\\bthund?ers?t(or|ro)ms?\\s?w[a-z0-9]*\\b"
evtypes <- mutate(evtypes, 
                  type.mapped=fixtext(type.mapped, pattern.flood, "flood"),
                  type.mapped=fixtext(type.mapped, pattern.freeze, "freeze"),
                  type.mapped=fixtext(type.mapped, pattern.wind, "wind"),
                  type.mapped=fixtext(type.mapped, pattern.thunderstormwind, 
                                      "thunderstorm wind"))


```

There are still `r length(unique(evtypes$type.mapped))` unique event types. Time constraints will require this to be set aside as future work. This will understate any sum-over-EVTYPE value because each variation will have its own category. This will definitely need to be revisited. 

## Storm effects on health
With a (relatively) clean list of event types, lets look at how each type can be categorized by effect on health. 

### Event Severity
Storm severity can be viewed as combination of fatilities and injuries incurred by an event. In this case we'll use fatalities + injuries as shown the the below plot.

```{r health.storm.results}
summarydata  <- stormdat %>%
    mutate(event=factor(EVTYPE)) %>%
    group_by(event) %>%
    summarize(Fatalities=sum(FATALITIES), Injuries=sum(INJURIES)) %>% 
    filter(Fatalities > 100, Injuries > 1000) %>%
    arrange(desc(Fatalities), desc(Injuries)) %>%
    mutate(event=factor(event, as.character(event)))
summarydata_melt <- melt(summarydata, "event")
ggplot(summarydata_melt, aes(event, value, fill=variable)) + 
    geom_bar(stat = "identity", position="dodge") + 
    labs(title="Worst Natural Events for Public Health") +
    labs(x="Natural Event") + 
    labs(y="Count of Fatalities and Injuries") + 
    theme(axis.text.x = element_text(angle = -30, hjust = 0))
```

## Storm effects on economy
Storms also effect the economy in terms of property and crop damage. 

### Event Severity
Event severity for damage is simply the total monetary damage to property and crops. 