---
title: "Reproducible Research: Storm Severity Analysis"
output: 
  html_document:
    keep_md: true
---

## Questions
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Project Setup

The libraries used in this analysis are:
```{r setup}
library(data.table)
library(plyr); library(dplyr)
library(ggplot2)
library(stringr)
```

## Loading and preprocessing

The data comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. An intermediate download location for this class is hosted on cloudfront:

```{r download, cache=TRUE}
setwd("~/../datascience/storm-severity")
dir.create("data")

dlURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filepath <- "data/StormData.csv.bz2"

# set browser to download file through IE protocol (for https)
setInternet2(use = TRUE)
download.file(dlURL, destfile = filepath, mode="wb")
```

```{r load, cache=TRUE}
stormdata <- read.csv(filepath, stringsAsFactors=FALSE);
```

To answer the two questions posed earlier we only need to care about the columns for `EVTYPE`, `FATALITIES`, `INJURIES`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG`, `CROPDMGEXP`, and `REMARKS`. 
```{r preprocess}
stormdat <- data.table(select(stormdata, EVTYPE, FATALITIES:CROPDMGEXP, REMARKS))
stormdat <- mutate(stormdat, EVTYPE=factor(EVTYPE), 
                   PROPDMGEXP=factor(PROPDMGEXP), 
                   CROPDMGEXP=factor(CROPDMGEXP))
dim(stormdat)
sapply(stormdat[1,], class)
head(stormdat)
```

## Any missing Values? 
```{r NAcount}
NA.count <- !complete.cases(select(stormdat, FATALITIES:PROPDMG, CROPDMG))
```
There are `r sum(NA.count)` rows with NA values.

## Managing EVTYPE variable
There are `r length(levels(stormdat$EVTYPE))` different event types in `EVTYPE`, which is much larger than the defined list in the [NOAA instruction sheet](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). Let's see if this can be cleaned up. 

First we exclude any entry which contains no adverse effects.
```{r remove.ineffective.rows}
# filter for adverse effects > 0
stormdat <- filter(stormdat, (FATALITIES+INJURIES+PROPDMG+CROPDMG) > 0)
dim(stormdat)
length(levels(stormdat$EVTYPE))
```
That didn't seem to have much effect, but at least we can work with fewer data.

Taking a look at values of `EVTYPE`:
```{r evtype.values}
head(unique(stormdat$EVTYPE), 15)
```
shows several spelling variations for the entry `THUNDERSTORM WINDS`. Also various hurricanes are named and labeled with variations such as `Hurricane Opal`, `Hurricane Erin`, `Hurricane Opal/High Winds`. 

The "real" event types are extracted from the table of contents of the NOAA instruction sheet.
```{r evtype.categorize}
# get event types from event types text file
evtypes.raw <- readLines("data/event types.txt", warn=F)

# extract 1st-level event types, excluding "...", page numbers and other data
#   "7.2 Words, (with-stuff/other stuff) "
evtypes.def <- str_extract_all(evtypes.raw, 
                               "^[0-9]+\\.[0-9]+\\s[A-Za-z\\s,/()-]{2,}")

# remove empty rows
evtypes.def <- unlist(evtypes.def[sapply(evtypes.def, length) > 0])

# remove leading numbers and trailing (<char>)
evtypes.def <- str_sub(evtypes.def, 
                       str_locate(evtypes.def, "[0-9] ")[,2]+1, 
                       str_locate(evtypes.def, " \\([A-Z]\\)")[,1]-1)
```

Custom-defined catagories for disaster will be used to categorize the `EVTYPE` column. 
## Storm effects on health

### Definition of storm severity

### Worst event type
```{r top.disasters}

```

## Storm effects on economy

### Definition of storm severity

### Worst event type